---
title: "Energy Trend Project"
author: "Bryan Depew"
format: 
  html:
    toc: true
    toc-location: left
    theme: cosmo
    code-fold: true
execute:
  warning: false
  message: false
  echo: true
---

##Mission Statement:

Clean and utilize U.S Energy Information Administration data to uncover and clearly communicate trends in energy production across natural gas, coal, petroleum/oil, and biomass/waste energy sectors. Apply findings to geosptaial state polygons (tigris package) to emphasize regional differences and fluctuations in energy use and production.



## Loading necessary packages and setting energy categories of interest
```{r}

library(tidyverse)
library(data.table)
library(tigris)
library(sf)
library(dplyr)
library(ggplot2)

options(tigris_use_cache = TRUE)

categories_of_interest <- c("SUN", "NG", "OG", "WAT", "BIT", "SUB", "WND", "RC", "WC", "DFO", "RFO", "WO", "JF", "PG", "PC", "NUC", "WDS", "WDL", "BLQ", "LFG", "MSB", "MSN", "OBG", "OBL", "OBS", "AB", "SLW", "TDF" )

solar_cats <- c("SUN")
wind_cats <- c("WND")
hydro_cats <- c("WAT")
nat_gas_cats<- c("NG", "OG")
coal_cats <- c("BIT", "SUB", "RC", "WC")
pet_oil_cats<- c("DFO", "RFO", "WO", "JF", "PG", "PC")
bio_waste_cats<- c("WDS", "WDL", "BLQ", "LFG", "MSB", "MSN", "OBG", "OBL", "OBS", "AB", "SLW", "TDF")





```



## Create and filter dataset of power plants and their energy sectors
```{r} 

plant_data_2023 <- read.csv("C:/Users/bgdep/Downloads/Energy Trend Project/List_of_plants_for_all_fuels_South_Atlantic_(total)_all_sectors.csv", skip = 4)%>%
  filter(Fuel.Types %in% categories_of_interest)

plant_data_2023 <- plant_data_2023%>%
  mutate(fuel_cat = recode(Fuel.Types,
                           "SUN" = "SOLAR",
                           "WND" = "WIND",
                           "WAT" = "HYDRO",
                           "NG" = "NATURAL GAS",
                           "OG" = "NATURAL GAS",
                           "BIT" = "COAL",
                           "SUB" = "COAL",
                           "RC" = "COAL",
                           "WC" = "COAL",
                           "DFO" = "PETROLEUM/OIL",
                           "RFO" = "PETROLEUM/OIL",
                           "WO" = "PETROLEUM/OIL",
                           "JF" = "PETROLEUM/OIL",
                           "PG" = "PETROLEUM/OIL",
                           "PC" = "PETROLEUM/OIL",
                           "WDS" = "BIOMASS/WASTE",
                           "WDL" = "BIOMASS/WASTE",
                           "BLQ" = "BIOMASS/WASTE", 
                           "LFG" = "BIOMASS/WASTE",
                           "MSB" = "BIOMASS/WASTE",
                           "MSN" = "BIOMASS/WASTE",
                           "OBG" = "BIOMASS/WASTE",
                           "OBL" = "BIOMASS/WASTE",
                           "OBS" = "BIOMASS/WASTE",
                           "AB" = "BIOMASS/WASTE",
                           "SLW" = "BIOMASS/WASTE",
                           "TDF" = "BIOMASS/WASTE"))%>%
                           select(c("State","Plant.Code", "fuel_cat"))
```


#Create, modify, and clean data set of states and their corresponding energy production by sector. Identify predominant fuel type by state and year.
```{r}

gen_data <- read.csv("C:/Users/bgdep/Downloads/Energy Trend Project/Net_generation_for_all_sectors.csv", skip = 4)
gen_data <- na.omit(gen_data)%>%
  mutate(STATEID = NA)%>%
  mutate(across(4:24, as.numeric))

for (i in 1:nrow(gen_data)) {
  key <- gen_data$source.key[i]
  pieces <- strsplit(key, "-")[[1]]
  gen_data$STATEID[i] <- pieces[2]

  first_part <- pieces[1]
  pieces2 <- strsplit(first_part, "\\.")[[1]]
  gen_data$FUEL[i] <- pieces2[3]
}

gen_categories_of_interest <- c("SUN", "TSN", "STH", "SPV", "DPV",
                                 "WND",
                                 "HYC", "HPS",
                                 "NG", "OOG",
                                 "COW",
                                 "PEL", "PC",
                                 "BIO", "WAS", "WWW")

gen_data <- gen_data %>%
  mutate(fuel_cat = recode(FUEL,
                           "SUN" = "SOLAR", "TSN" = "SOLAR", "STH" = "SOLAR", "SPV" = "SOLAR", "DPV" = "SOLAR",
                           "WND" = "WIND",
                           "HYC" = "HYDRO", "HPS" = "HYDRO",
                           "NG" = "NATURAL GAS",
                           "COW" = "COAL",
                           "PEL" = "PETROLEUM/OIL", "PC" = "PETROLEUM/OIL",
                           "BIO" = "BIOMASS/WASTE"))

gen_data <- gen_data %>%
  filter(FUEL %in% gen_categories_of_interest)

south_atlantic_states <- states() %>%
  filter(STUSPS %in% unique(gen_data$STATEID))
##Pivot longer to easily utilize data
gen_long <- gen_data %>%
  pivot_longer(
    cols = starts_with("X"),
    names_to = "year",
    values_to = "generation"
  ) %>%
  mutate(year = as.numeric(gsub("X", "", year)))

## find the dominant fuel type per state, per year
dominant_fuel_by_year <- gen_long %>%
  group_by(STATEID, year) %>%
  slice_max(generation, n = 1, with_ties = FALSE) %>%
  select(STATEID, year, FUEL)

```


##Create geospatial map of findings
```{r}

map_data <- south_atlantic_states %>%
  left_join(dominant_fuel_by_year, by = c("STUSPS" = "STATEID"))%>%
  mutate("Predominant Fuel Type" = recode(FUEL,
                            "COW" = "Coal",
                            "NG" = "Natural Gas",
                            "PEL" = "Petroleum/Oil",
                            "BIO" = "Biomass/Waste"))

map_data_subset <- map_data %>%
  filter(year %in% c(2005, 2010, 2015, 2020, 2025))

map <- ggplot(map_data_subset) +
  geom_sf(aes(fill = `Predominant Fuel Type`)) +
  facet_wrap(~ year) +
  theme_void()

plot(map)
```

##Communicate trend of energy production over time by sector

```{r}
coal_plot <- gen_long %>%
  filter(fuel_cat == "COAL") %>%
  ggplot(aes(x = year, y = generation, color = STATEID, group = STATEID)) +
  geom_line()+
  labs(title = "Coal Energy Generation by State (in thousand megawatthours)", y= "Energy Generation", x = "Year" )

coal_plot

nat_gas_plot <- gen_long %>%
  filter(fuel_cat == "NATURAL GAS") %>%
  ggplot(aes(x = year, y = generation, color = STATEID, group = STATEID)) +
  geom_line()+
  labs(title = "Natural Gas Energy Generation by State (in thousand megawatthours)", y= "Energy Generation", x = "Year" )

nat_gas_plot

pet_oil_plot <- gen_long %>%
  filter(fuel_cat == "PETROLEUM/OIL") %>%
  ggplot(aes(x = year, y = generation, color = STATEID, group = STATEID)) +
  geom_line()+
  labs(title = "Petroleum/Oil Energy Generation by State (in thousand megawatthours)", y= "Energy Generation", x = "Year" )

pet_oil_plot

biomass_plot <- gen_long %>%
  filter(fuel_cat == "BIOMASS/WASTE") %>%
  ggplot(aes(x = year, y = generation, color = STATEID, group = STATEID)) +
  geom_line()+
  labs(title = "Biomass/Waste Energy Generation by State (in thousand megawatthours)", y= "Energy Generation", x = "Year" )

biomass_plot


```
The `echo: false` option disables the printing of code (only output is displayed).
